---
title: "02 Preprocess Flight History"
author: "Nils Hellwig"
date: "5/16/2023"
output: html_document
---

## Load Packages

```{r}
library(rvest)
library(tidyverse)
library(stringr)
library(xml2)
source("date_string_formatter.R")
library(lubridate)
```

## Define Settings

```{r}
folder_path_departures <- "datasets/raw_departures_dataset"
```

## Preprocessing of departure flights

```{r}
# Load the filepaths of all the .html-documents
html_documents <- list.files(path = folder_path_departures, pattern = "\\.html$", full.names = TRUE)
head(html_documents)
```

```{r}
departure_data <- data.frame(
  departure_scheduled_datetime_local = character(),
  departure_scheduled_time_local_timezone = character(),
  departure_scheduled_datetime_utc = character(),
  flight_status = character(),
  airline_name = character(),
  flight_number = character(),
  flight_number_IATA = character(),
  airline_code = character(),
  airline_code_IATA = character(),
  airline_country = character(),
  destination_country = character(),
  destination_name = character(),
  destination_IATA = character(),
  destination_ICAO = character(),
  departure_actual_datetime_local = character(),
  delay_departure_min = numeric(),
  delay_arrival_min = numeric(),
  actual_arrival_datetime_local = character(),
  flight_duration_time_min = numeric()
)
```

```{r}
# Load each .html-document
for (i in seq_along(html_documents)) {
  doc_path <- html_documents[i]
  doc <- read_html(doc_path)
  
  extracted_data <- doc %>%
    html_nodes("tbody tr.bg-gray-50, tbody tr.bg-white")
  
  for (flight_element in extracted_data) {
      # let's start by storing all child elements of a flight's flight_element 
      child_elements_txt <- flight_element %>%
        html_children() %>%
        html_text() %>%
        trimws() # remove whitespaces
      
      flag_imgs <- flight_element %>%
         html_nodes("img[alt*='Flag for']")

      # It happens that the airline can not be assigned to a country so that there's no flag icon.
      # Example: Flight to Liege (Belgium), see here: https://www.flightera.net/en/airport/Munich/EDDM/departure/2017-12-13%2022_00?
      if (length(flag_imgs) == 1) {
         airline_country <- NULL
         destination_country <- str_extract(html_attr(flag_imgs[1], "alt"), "(?<=Flag for ).*")
      } else {
         airline_country <- str_extract(html_attr(flag_imgs[1], "alt"), "(?<=Flag for ).*")
         destination_country <- str_extract(html_attr(flag_imgs[2], "alt"), "(?<=Flag for ).*")
      }
     
      # There are either one or two flight numbers. The flight number is located before the first two \n\n inside child_elements_txt[2]
      flight_number <- gsub("\\n\\n.*", "", child_elements_txt[2])
      flight_number <- str_extract_all(flight_number, "[A-Z\\d]+")
      
      airline <- strsplit(child_elements_txt[2], "\n")[[1]]
      airline <- tail(airline, n = 1)
    
      airline <- strsplit(airline, "/")[[1]]
      airline_code <- tail(strsplit(airline[1], " ")[[1]], n = 1)
      airline_code_IATA <- airline[2]
      airline_name <- sub("\\s\\S+$", "", airline)[1]
      
      
      departure_scheduled_date_local <- convert_date_string(strsplit(child_elements_txt[1], "\n")[[1]][1])
      flight_status <- tail(gsub("\\s+", "",strsplit(child_elements_txt[1], "\n")[[1]]), n = 1)
 
           
      # Although we know that the local timezone of munich is CEST, I decided to extract that data too because 
      # one might want to use the script for another airport
      if (year(as.Date(departure_scheduled_date_local))==2023) {
        departure_scheduled_time_local <- strsplit(child_elements_txt[1], "\n")[[1]][7]
        departure_scheduled_time_local_timezone <- strsplit(child_elements_txt[1], "\n")[[1]][8]
      } else {
        departure_scheduled_time_local <- strsplit(child_elements_txt[1], "\n")[[1]][8]
        departure_scheduled_time_local_timezone <- strsplit(child_elements_txt[1], "\n")[[1]][9]
      }

      departure_scheduled_datetime_local <- paste(departure_scheduled_date_local, departure_scheduled_time_local)
      departure_scheduled_datetime_utc <- format(with_tz(ymd_hm(departure_scheduled_datetime_local, tz = "CEST"), "UTC"), "%Y-%m-%d %H:%M")
     
      destination_name <- sub(" \\(.*", "", child_elements_txt[3])
      
      # str_match returns the second column of these matrices
      flight_code_IATA <- str_match(child_elements_txt[3], "\\(([A-Z]{3}) /")[, 2]
      flight_code_ICAO <- str_match(child_elements_txt[3], " / ([A-Z]{4})\\)")[, 2]
      
      if (flight_status == "Landed") {
        delay_departure_split <- strsplit(child_elements_txt[5], "\n")[[1]]
        delay_departure_min <- extract_minutes_from_delay_string(delay_departure_split)
        departure_actual_datetime_local <- format(ymd_hm(departure_scheduled_datetime_local) + minutes(delay_departure_min), "%Y-%m-%d %H:%M")
        delay_arrival_split <- strsplit(child_elements_txt[6], "\n")[[1]]
        delay_arrival_min <- extract_minutes_from_delay_string(delay_arrival_split)
        
        flight_duration_time_min <- extract_minutes_from_delay_string(child_elements_txt[7])
        actual_arrival_datetime_local <- format(ymd_hm(departure_scheduled_datetime_local) + minutes(flight_duration_time_min + delay_departure_min), "%Y-%m-%d %H:%M")
      }
      
      
      flight_number_IATA <- flight_number[[1]][2]
      flight_number <- flight_number[[1]][1]
      
      # Finally, we can add all the extracted information regarding a specific flight to our dataframe
      departure_data <- departure_data %>%
        add_row(
          departure_scheduled_datetime_local = departure_scheduled_datetime_local,
          departure_scheduled_time_local_timezone = departure_scheduled_time_local_timezone,
          departure_scheduled_datetime_utc = departure_scheduled_datetime_utc,
          airline_code = airline_code,
          airline_code_IATA = airline_code_IATA,
          airline_name = airline_name,
          airline_country = airline_country,
          flight_number = flight_number[[1]][1],
          flight_number_IATA = flight_number_IATA,
          flight_status = flight_status,
          departure_actual_datetime_local = departure_actual_datetime_local,
          destination_name = destination_name,
          destination_IATA = flight_code_IATA,
          destination_ICAO = flight_code_ICAO,
          destination_country = destination_country,
          delay_departure_min = delay_departure_min,
          delay_arrival_min = delay_arrival_min,
          actual_arrival_datetime_local = actual_arrival_datetime_local,
          flight_duration_time_min = flight_duration_time_min
        )
  }
}
```

If there are not at least 30 flights available for an hour on flightera.com, more than one randomly selected hour may be crawled, in rare cases duplicates may occur. This is because if there are less than 30 flights available, the flights for the hours after the randomly selected hour are returned - to get 30 flights in each case.

```{r}
duplicate_rows <- departure_data[duplicated(departure_data), ]
departure_data <- distinct(departure_data)
```

```{r}
# For better data handling, I´ll create separate columns for month, year and hour of the planned departures.
departure_data$departure_scheduled_datetime_local <- as.POSIXct(departure_data$departure_scheduled_datetime_local, format = "%Y-%m-%d %H:%M")
departure_data$departure_scheduled_month_year_local <- format(departure_data$departure_scheduled_datetime_local, "%Y-%m")
departure_data$departure_scheduled_month_local <- format(departure_data$departure_scheduled_datetime_local, "%m")
departure_data$departure_scheduled_year_local <- format(departure_data$departure_scheduled_datetime_local, "%Y")
departure_data$departure_scheduled_hour_local <- format(departure_data$departure_scheduled_datetime_local, format = "%H")

# Setting the language with Sys.setenv is relevant to show the delays correctly depending on the week (names should be in English in the plot)
Sys.setlocale("LC_TIME", "C")
weekday_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
departure_data$departure_scheduled_weekday_local <- factor(weekdays(departure_data$departure_scheduled_datetime_local), levels = weekday_order)
```


```{r}
departure_data
```


```{r}
write.csv(departure_data, file = "datasets/departure_dataset.csv", row.names = FALSE)
```
