# Notebook: Create Dataframe with Weather data

This notebook is used to compile the weather data downloaded as .html files into a dataframe. This is a particular challenge because the weather data in a .html file is inside a `<script/>` tag. The data is stored in a JavaScript object.

## Load Packages

```{r}
library(rvest)
library(tidyverse)
library(stringr)
library(xml2)
library(jsonlite)
```

## Define Settings

```{r}
folder_path_weather_munich <- "../datasets/raw_weather_munich_dataset"
```

## Code

### Preprocessing of departure flights

```{r}
# Load the filepaths of all the .html-documents
html_documents <- list.files(path = folder_path_weather_munich, pattern = "\\.html$", full.names = TRUE)
head(html_documents)
```

### Init empty dataframe to store weather data

```{r}
weather_data <- data.frame(
  ds = character(),
  icon = numeric(),
  desc = character(),
  temp = numeric(),
  templow = numeric(),
  baro = numeric(),
  wind = numeric(),
  wd = numeric(),
  hum = numeric()
)
```

### Iterate through all html-documents and store data to .html-file

The weather data is always in the second `<script>`-element within the element of class headline-banner__wrap. It looks like this:

```
<script type="text/javascript">
      var data = { 
           ...
           detail: [
               {
                      hl: true,
                      hls: "Fr, 1. Dez",
                      hlsh: "1. Dez",
                      date: 1512108e6,
                      ts: "06:00",
                      ds: "Freitag, 1. Dezember 2017, 06:00 — 12:00",
                      icon: 6,
                      desc: "Partly sunny.",
                      temp: 0,
                      templow: -6,
                      baro: 1010,
                      wind: 5,
                      wd: 250,
                      hum: 95,
                }, ...
           ],
           ...
      }
<script/>
````

There is always weather data for the time between 00:00-6:00, 6:00-12:00, 12:00-18:00 and 18:00-24:00.

```{r}
for (html_path in html_documents) {
  html <- read_html(html_path)
  script_tag_inner <- html %>% html_nodes(".headline-banner__wrap script:nth-child(2)") %>% html_text()
  
  # This expression is used to select the inner of the script tag. The weather data is stored in a JavaScript variable inside a <script>-tag
  script_tag_inner <- gsub(".*detail(.+)\\]", "\\1", script_tag_inner)
  script_tag_inner <- gsub("\n", "", script_tag_inner)
  script_tag_inner <- gsub('\\"', "'", script_tag_inner)
  
  extracted_strings <- str_extract_all(script_tag_inner, "\\{([^\\{\\}]*)\\}")[[1]]

  for (weather_data_string in extracted_strings) {
    weather_row <- data.frame(
      ds = NA,
      icon = NA,
      desc = NA,
      temp = NA,
      templow = NA,
      baro = NA,
      wind = NA,
      wd = NA,
      hum = NA
    )
    
    matches <- str_extract_all(weather_data_string, "'(ds|desc)':('([^']*)')")
    
    # First, extract the textual data that is relevant for me. 
    # This is the description (for example "Partly sunny.") and ds (date string, for example "Freitag, 1. Dezember 2017, 06:00 — 12:00")
    for (match in matches[[1]]) {
      key <- str_extract(match, "ds|desc")
      value <- str_extract(match, "(:'([^']*)')")
      # this expression is used to remove the "'" at the beginning of a string stored in the JavaScript object as well as the colon before the value
      value <- str_replace(value, ":'", "")
      # this expression is used to remove the "'"-sign at the end of a string stored in the JavaScript object
      value <- str_replace(value, "'$", "")
      weather_row[[key]] <- value
    }
    # Sometimes, there are objects listed within the detail array without actual weather data for a specific point in time. 
    # for example: "{'offset':0,'scale':1}", which is useless for my purpose.
    # in order to obtain only relevant data, it is checked whether there is the column "ds" or not
    if (any(!is.na(weather_row$ds)) && weather_row["desc"] != "") {
       matches <- str_extract_all(weather_data_string, "'(icon|temp|templow|baro|wind|wd|hum)':-?[0-9]+")
       # Especially for the time before 2020, there are some periods for which no historical weather data are available.
       # It does not make sense to extract weather data for this case, as it is not stored in the object.
       if (weather_row$desc != "No weather data available") {
            for (match in matches[[1]]) {
               key <- str_extract(match, "'icon'|'temp'|'templow'|'baro'|'wind'|'wd'|'hum'")
               key <- str_replace_all(key, "'", "")
               value <- str_extract(match, "(:-?[0-9]+)")
               value <- str_replace(value, "\\:", "")
               weather_row[[key]] <- value
            } 
       }
       weather_data <- rbind(weather_data, weather_row)
    }
  }
}
```

### Further Preprocessing

```{r}
convert_datetime <- function(string) {
  date_time <- strptime(string, "%A, %d. %B %Y, %H:%M")
  return(format(date_time, "%Y-%m-%d %H:%M:%S"))
}

weather_data$ds <- sapply(weather_data$ds, convert_datetime)
head(weather_data)
```

```{r}
# Store  date and time in a separate column
weather_data$ds <- as.POSIXct(weather_data$ds)
weather_data$date <- format(weather_data$ds, "%Y-%m-%d")
weather_data$six_hours_starting_from <- format(weather_data$ds, "%H:%M")
weather_data <- subset(weather_data, select = -ds)
head(weather_data)
```

```{r}
# Make sure that the dataframe is sorted by date and time
weather_data <- weather_data %>%
  arrange(date, six_hours_starting_from)
head(weather_data)
```

```{r}
# Add better names for columns
names(weather_data)[names(weather_data) == 'desc'] <- 'weather_description'
names(weather_data)[names(weather_data) == 'temp'] <- 'temperature_celsius'
names(weather_data)[names(weather_data) == 'templow'] <- 'temperature_celsius_low'
names(weather_data)[names(weather_data) == 'wd'] <- 'wind_direction'
names(weather_data)[names(weather_data) == 'wind'] <- 'wind_speed_km'
names(weather_data)[names(weather_data) == 'hum'] <- 'humidity'
# Id of the weather icon. Example: an entry in weather_data with id=19 can be found here: https://c.tadst.com/gfx/w/svg/wt-19.svg (rain)
names(weather_data)[names(weather_data) == 'icon'] <- 'weather_icon'
head(weather_data)
```

```{r}
write.csv(weather_data, file = "../datasets/weather_munich_dataset.csv", row.names = FALSE)
```
